Reproducible development environments with rix

### 2024-02-09 paper.qmd
Update paper.qmd

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which is available for all major operating systems and focuses
on installing and building software in a reproducible manner. This means that
this single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg}
and even a build automation tool such as [Make]{.pkg}. To make
[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
[rix]{.pkg} package which we will present in this article.

@@

[Nix]{.pkg} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using [Nix]{.proglang}, it is possible to install not only
[R]{.proglang}, but also all the packages required for a project. The reason why
one should use [Nix]{.proglang} to install [R]{.proglang} packages, and not use
the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

@@

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`. This is
not the only reason [Nix]{.pkg} is very useful for reproducibility, but the
[Nix]{.pkg} package manager itself was designed with reproducibility in
mind. [Nix]{.pkg} is functional package manager, meaning that it uses
concepts from functional programming to build software: no-side effects or
global variables are allowed, and the output is always the same if given the
same inputs, regardless of where or when a package gets built. [Nix]{.pkg} 
also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
to be installed at the same time on the same system. Each variant of a package
has a unique identifier, so that it is possible to have two versions of [R]{.proglang}
installed and use the correct one where adequate.
We refer to @dolstra2004nix for a more technical discussion on [Nix]{.pkg}.

With Nix, it is essentially possible to replace [renv]{.pkg} and
[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
`requirements.txt` files which list package dependencies for [Python]{.proglang}
projects. It is also possible to build multi-language environments, containing R
and Python, a LaTeX distribution, and even install one's favorite text editor to
edit the project's source code. Essentially, one can define a complete
reproducible environment and use it to work on a project interactively, or use
that environment non-interactively to run an analysis sometime in the future.

@@

[Nix]{.pkg} expressions are written in the [Nix]{.proglang} programming language,
which is a purely functional programming language. Here is a simple expression
that builds a shell environment containing version 4.3.1 of [R]{.proglang}:

@@

To define a variable in the [Nix]{.proglang} language, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github repository, and in particular the commit starting
with `976fa336` is going to be used. It should be noted that is also possible to
use a personal fork of `nixpkgs` instead of the official repository.
`system_pkgs` lists the packages that we wish to install and use with this
environment, in this example only the [R]{.proglang} programming language will
be installed, alongside all the required dependencies to actually install and
run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell,
[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
startup options).

This expression can be saved in a file called `default.nix`, and the shell can
be built using the `nix-build` command on a system where [Nix]{.proglang} is
available.^[To install [Nix]{.proglang} we recommend the installer from
Determinate systems which can be found on the following link:
\url{https://determinate.systems/posts/determinate-nix-installer}.] Once
`nix-build` is done building the development environment according to the
specifications declared in the `default.nix` file, the user can enter the
development shell using `nix-shell`. The user will thus be dropped into an
interactive shell with all the packages declared in the `default.nix` file and
can now use this shell for work. This is similar to activating environments in
the [Python]{.proglang} ecosystem.

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite appealing. Because we believe that lowering the cost of
writing [Nix]{.proglang} expressions would lead to a wider adoption of the
[Nix]{.proglang} package manager for reproducibility, we wrote [rix]{.pkg}
which comes with the `rix()` function that helps generate such expressions quite
easily. For example, here is how one could use [rix]{.pkg} to generate a
`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
expression here because of space):

@@

rix(r_ver = "4.3.1",
    r_pkgs = c("dplyr", "chronicler"),
    project_path = ".",
    overwrite = TRUE,
    print = TRUE)
```

Here, users do not need to look for the right commit of `nixpkgs`: it is enough
to specify the version of [R]{.proglang} they wish to use and the packages that
were current at the time will get used. But [rix]{.pkg} is quite flexible and
makes it easy to install specific versions of packages as well (which will get
installed from the CRAN archives) as well as packages hosted on Github and
\LaTeX packages as well:

@@

[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
system. This is quite useful if one wants to generate an expression and use that
expression in a CI/CD environment such as Github Actions. For example, the
repository that contains the source code to this
article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
compilation. Essentially, each time one of the authors pushed to the master
branch, a Github Actions runner would be started, [Nix]{.proglang} would get
installed on said runner, and an environment containing the right packages was
generated from a `default.nix` file that is hosted on the same repository. The
paper would then get compiled in this reproducible environment.

@@

We recommend `path_subshell` to be a subfolder of the project. Using `rix()` we
generate a `default.nix` file that will provide a shell with R version 3.5.3
when built. We then use `rix_init()`, which generates an `.Rprofile` file in the
same subfolder as where the subshell's `default.nix` is located. This
`.Rprofile` file ensures that the R session that will be spawned from this
subshell does not get any interference from the main session it gets called
from:

@@

The `expr` argument of `with_nix()` can be arbitrarly complex, but must be
wrapped inside a `function()` call. The value of `out_subshell` is `r out_subshell`,
as is indeed different than the one generated from a recent version of R.

@@

>>>>

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which is available for all major operating systems and focuses
on installing and building software in a reproducible manner. This means that
this single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg}
and even a build automation tool such as [Make]{.pkg}. To make
[Nix]{.pkg} more accessible to [R]{.proglang} programmers, we developed the
[rix]{.pkg} package which we will present in this article.

@@

[Nix]{.pkg} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using [Nix]{.pkg}, it is possible to install not only
[R]{.proglang}, but also all the packages required for a project. The reason why
one should use [Nix]{.pkg} to install [R]{.proglang} packages, and not use
the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

@@

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`. This is
not the only reason [Nix]{.pkg} is very useful for reproducibility, but the
[Nix]{.pkg} package manager itself was designed with reproducibility in
mind. [Nix]{.pkg} is functional package manager, meaning that it uses
concepts from functional programming to build software: no-side effects or
global variables are allowed, and the output is always the same if given the
same inputs, regardless of where or when a package gets built (an important
side-note is in order here: while the functional aspect of [Nix]{.pkg} is quite
useful to ensure reproducibility, it can lead to situations where writing the
required [Nix]{.proglang} expression to build it can be much more difficult than usual: this is especially true for
packages that must download assets off the internet during installation, such as 
several BioConductor packages like [musData]{.pkg} for example. But this is only
a concern for the packagers, not users of the package that simply wish to install
and use the package). [Nix]{.pkg} also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
to be installed at the same time on the same system. Each variant of a package
has a unique identifier, so that it is possible to have two versions of [R]{.proglang}
installed and use the correct one where adequate.
We refer to @dolstra2004nix for a more technical discussion on [Nix]{.pkg}.

With Nix, it is essentially possible to replace [renv]{.pkg} and
[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
`requirements.txt` files which list package dependencies for [Python]{.proglang}
projects. It is also possible to build multi-language environments, containing R
and Python, a LaTeX distribution, and even install one's favorite text editor to
edit the project's source code. Essentially, one can define a complete, project-specific,
and reproducible environment and use it interactively, or non-interactively. It will
always be possible to rebuild this environmen in the future (as long as the `nixpkgs`
GitHub repository is online) to run an analysis sometime in the future.

@@

As already stated, [Nix]{.pkg} expressions are written in the [Nix]{.proglang} programming language,
which is a purely functional programming language. Here is a simple expression
that builds a shell environment containing version 4.3.1 of [R]{.proglang}:

@@

To define a variable in the [Nix]{.proglang} language, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github repository, and in particular the commit starting
with `976fa336` is going to be used (it should be noted that is also possible to
use a personal fork of `nixpkgs` instead of the official repository).
`system_pkgs` lists the packages that we wish to install and use with this
environment; in this example only the [R]{.proglang} programming language will
be installed, alongside all of its dependencies, and all of the dependencies’
dependencies. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell,
[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
startup options).

This expression can be saved in a file called `default.nix`, and the shell can
be built using the `nix-build` command on a system where [Nix]{.proglang} is
available.^[To install [Nix]{.proglang} we recommend the installer from
Determinate systems which can be found on the following link:
\url{https://determinate.systems/posts/determinate-nix-installer}.] Once
`nix-build` is done building the development environment according to the
specifications declared in the `default.nix` file, the user can enter the
development shell using `nix-shell`. The user will thus be dropped into an
interactive shell with all the packages declared in the `default.nix` file and
can now use this shell for work. This is similar to activating virtual environments in
the [Python]{.proglang} ecosystem.

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite appealing. Because we believe that lowering the cost of
writing [Nix]{.proglang} expressions would lead to a wider adoption of the
[Nix]{.pkg} package manager for reproducibility, we wrote [rix]{.pkg}. [rix]{.pkg}
comes with the `rix()` function that helps generate such expressions quite
easily. For example, here is how one could use [rix]{.pkg} to generate a
`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
expression here because of space):

@@

rix(r_ver = "4.3.1",
    r_pkgs = c("dplyr", "chronicler"),
    project_path = ".",
    overwrite = TRUE)
```

Here, users do not need to look for the right commit of `nixpkgs`: it is enough
to specify the version of [R]{.proglang} they wish to use and the right
commit will be selected (a lookup table linking versions of R and `nixpkgs` commits
is included in the package). The versions of [R]{.proglang} packages that
were current at the time will get installed in that environment. But [rix]{.pkg} is quite flexible and
makes it easy to install specific versions of packages as well (which will get
installed from CRAN) as well as packages hosted on Github and
\LaTeX packages as well:

@@

[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
system. This is quite useful if one wants to generate an expression and use that
expression in a CI/CD environment such as Github Actions. For example, the
repository that contains the source code to this
article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
compilation. Essentially, each time one of the authors pushed to the master
branch, a Github Actions runner would be started, [Nix]{.proglang} would get
installed on said runner, and an environment containing the right packages was
generated from a `default.nix` file that is hosted on the same repository. The
paper would then get compiled using [Quarto]{.pkg} in this reproducible environment.

Instead of first dropping into a [Nix]{.pkg} shell it is also possible to directly
run a program from the environment:

```
cd /absolute/path/to/project/ && nix-shell default.nix --run "Rscript analysis.R"
```

This will run `Rscript` and execute the `analysis.R` script (which in this example
should be in the same directory as `default.nix`).

@@

We recommend `path_subshell` to be a subfolder of the main project. Using `rix()` we
generate a `default.nix` file that will provide a shell with R version 3.5.3
when built. We then use `rix_init()`, which generates an `.Rprofile` file in the
same subfolder as where the subshell's `default.nix` is located. This
`.Rprofile` file ensures that the R session that will be spawned from this
subshell does not get any interference from the main session it gets called
from:

@@

The `expr` argument of `with_nix()` can be arbitrarly complex, but must be
wrapped inside a `function()` call. The value of `out_subshell` is `r out_subshell`,
and is indeed different than the one generated from a recent version of R.

@@

## Conclusion
